A comparison on R tools for nonlinear least squares modeling

### 2023-03-14 RefactoringNLS/ComparingNLStools.Rmd
Edits

`nls2` (@nls2-22) -- Nonlinear least squares by brute force has some similar
motivations, but a longer history and using `nls()` within multiple trials.

`nlstac` (@m-nlstac) -- Set of functions implementing the algorithm described in Fernandez 
    Torvisco et al. (2018) for fitting separable nonlinear regression curves. The special 
    class of problem for which this package is intended is an important and difficult one.
    No vignette is provided, unfortunately.

@@

$$   Asym =  b_1 = 100 * c_1 $$
$$ exp(xmid/scal)  =  b_2 = 10 * c_2 $$
$$ 1/scal  =  b_3 $$
To allow for simpler discussion, let us say that the parameters for a vector
$p$ and the model function is called $model(p)$. The residuals can be written
either as

@@

since when squared the loss function has an identical value. In both cases, the
residual vector is a function of $p$. The second form
has a minor advantage in avoiding a potential sign error 
if we need to differentiate to get gradients etc.

@@

The general approach of nonlinear optimizers, of which nonlinear least squares
solvers are usually a subset, is to try to reduce the loss function (or objective
function) defined by the sum of squares of the residuals ($y - model()$) by 
altering the model parameters from some starting guess. In all general 
methods of which we are aware, this is an iterative process.

@@

$$ J' J \delta = - J' r$$
though we can avoid some loss of accuracy by NOT forming the inner product matrix
$J' J$ by solving the linear least squares matrix problem

@@

The suggestion is that $D$ be replaced by a multiple of the unit matrix or else
a multiple of the diagonal part of $J' J$. For machines with exceptionally low
precision arithmetic found on some minicomputers in the 1970s, one of us suggested
a linear combination of these ideas, as with low precision, some elements of 
$J' J$ could underflow to zero (@jn77ima). We note that the various choices for
$D$, especially the scaling multiple, as well as possible line search along the
direction $\delta$ rather than a unit step (@Hartley1961), 
give rise to a panoply of variant
algorithms. "Marquardt's method" is a family of methods. Fortunately, most choices
work well.

@@

One of the useful features of `nls()` is the possibility of a `selfStart` model,
where starting parameter values are not required. We discuss aspects of this
feature at several points below. However, in the event that a `selfStart` model 
is not available, `nls()` sets all the starting parameters 
to 1. This is, in our view, tolerable, but could be improved by using a set of values
that are all slightly different, which, in the case of the example 
model $y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)$
would avoid a singular Jacobian because $b$ and $d$ were equal in value. 
Program modifications to give a sequence like  1.0, 1.1, 1.2, 1.3 for the four 
parameters are fairly obvious. However, users may want to avoid equal parameter
starts in situations like this when setting starting values.

@@

Using a formula specification was one of the principal advances made
when `nls()` became available in S sometime in the 1980s. It uses a
Gauss-Newton (i.e., unstabilized) iteration with step reduction line
search. This works very efficiently as long as $J$ is not ill-conditioned.
In a one-parameter problem with ill-conditioning, think of $J$ as a 
very small number so that we need to watch for a zero-divide or 0/0 
situation in computing $\delta$.
The following show `nls()` does poorly on this example problem. Note that `nlsr`
offers function `pnlm0()` to give a 1-line display of class nls method results,
and `pshort()` for `nlsr`.

@@

The singular values displayed by `print.nlsr()` (the internal name for the 
adaptation of the generic `print()`) are displayed in a column to the right 
of the coefficient  and standard error display, but are NOT specific to the 
parameters. Their position is purely for efficient use of page space. The 
most common use of the singular values is to gauge how "nearly singular" the 
Jacobian is at the solution, and the ratio of the largest to smallest of the 
singular values is a simple but effective measure. In the above example, we note 
that the scaled logistic has the smallest ratio.

The results from `nlsLM` and `gsl_nls` for the transformed model have a very 
large sum of squares, which may suggest that these program have failed. 
Since `nls()`, `nlsLM()`, and `gsl_nls()` do not 
offer the singular values, we need to extract the Jacobian and compute its
singular values. The following script shows how to do this, using what is called
the `gradient` element in the returned solution for these solvers.

@@

More general termination tests can be used. `nlsr` a **small sum of 
squares** test (**smallsstest**) that compares the latest evaluated sum of squared
(weighted) residuals to `e4` times the initial sum of squares, where
`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.

@@

Using the "gradient" attribute of the output of the Jacobian function to hold
this matrix lets us embed this attribute of the **residual** function as well,
so that the call to `nlfb()` can be made with the same name used for 
both residual and Jacobian function arguments. This programming trick saves a 
lot of trouble for the package developer, but it can be a nuisance for users 
trying to understand the code.

As far as we can understand the logic in `nls()`, the Jacobian computation during
parameter estimation is carried out entirely within the called C-language program,
and the R code function `numericDeriv()`, part of `./src/library/stats/R/nls.R`
in the R distribution source code. This is used to provide Jacobian information in 
the `nlsModel()` and `nlsModel.plinear()` functions, which are **not** exported for 
general use. `gsl_nls()` also appears to use `numericDeriv()`.

@@

`minpack.lm::nlsLM()` invokes `numericDeriv()` in its local 
version of `nlsModel()`, but it appears to use an internal approximate Jacobian 
code from 
the original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
can lead to different behaviour, though in our experience the differences are usually
very minor.

Minor numerical differences can give annoying results with ill-conditioned
problems.

@@

`nls()` has an option `algorithm="plinear"` that allows some partially linear
models to be solved. The other tools, as far as we are aware, do not offer any
such capability, with no external package providing it. 
The feature in `nls()` has, however, some awkwardness.
In particular, we would like to specify a model to a solver consistently
across solver methods. Unfortunately, within `nls()` itself we must use
different specifications with different `algorithm` options.

The nonlinear model specifications are, of course, a development of linear ones.
Unfortunately, the explicit model `y ~ a * x + b` does not work with the linear
modeling function `lm()`, which requires this model to be specified as `y ~ x`.
Within `nls()`, consider the following FOUR different specifications for the same 
problem, plus an intuitive choice, labelled `fm2a`, that does NOT work. 
In this failed attempt, putting the `Asym` parameter in the model causes the
`plinear` algorithm
to try to add another term to the model. We believe this is unfortunate, and would
like to see a consistent syntax. At the time of writing, we do
not foresee a resolution for this issue. In the example, we have NOT evaluated
the commands to save space.

>>>>

`nls2` (@nls2-22) -- Nonlinear least squares by brute force has some similar
motivations, but uses `nls()` within multiple trials. The package and its
author have a long history in the R community.

`nlstac` (@m-nlstac) -- A set of functions implementing the algorithm described in Fernandez 
    Torvisco et al. (2018) for fitting separable nonlinear regression curves. The special 
    class of problem for which this package is intended is an important and difficult one.
    No vignette is provided, unfortunately.

@@

$$   Asym =  b_1 = 100 * c_1 $$
$$ exp(xmid/scal)  =  b_2 = 10 * c_2 $$
$$ 1/scal  =  b_3 $$
To allow for simpler discussion, let us say that the parameters form a (named) vector
$p$ and the model function is called $model(p)$. The residuals can be written
either as

@@

since the sumsquares loss function has an identical value. In both cases, the
residual vector is a function of $p$. The second form
has a minor advantage in avoiding a potential sign error 
if we need to differentiate to get gradients etc.

@@

The general approach of nonlinear optimizers, of which nonlinear least squares
solvers are usually a subset, is to try to reduce the loss function (or objective
function) defined by the sum of squares of the residuals by 
altering the model parameters from some starting guess. In all general 
methods of which we are aware, this is an iterative process.

@@

$$ J' J \delta = - J' r$$
though we can avoid some loss of accuracy by NOT forming the inner product matrix
$J' J$ and solving the linear least squares matrix problem

@@

The suggestion is that $D$ be replaced by a multiple of the unit matrix or else
a multiple of the diagonal part of $J' J$. For machines with exceptionally low
precision arithmetic found on some minicomputers in the 1970s, one of us suggested
a linear combination of these ideas. In low precision, some elements of 
$J' J$ could underflow to zero (@jn77ima). We note that the various choices for
$D$, especially the scaling multiple, as well as possible line search along the
direction $\delta$ rather than a unit step (@Hartley1961), 
give rise to a panoply of variant
algorithms. "Marquardt's method" is a family of methods. Fortunately, most choices
work well.

@@

One of the useful features of `nls()` is the possibility of a `selfStart` model,
where starting parameter values are not required. We discuss aspects of this
feature at several points below. However, in the event that a `selfStart` model 
is not available, `nls()` sets all the starting parameters 
to 1. This is, in our view, tolerable, but could be improved by using a set of values
that are all slightly different, which, in the case of the example 
model $y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)$
would avoid a singular Jacobian because $b$ and $d$ were equal in value. 
Program modifications to give a sequence like  1.0, 1.1, 1.2, 1.3 for the four 
parameters are fairly obvious.

@@

Using a formula specification was one of the principal advances made
when `nls()` became available in S sometime in the 1980s. It uses a
Gauss-Newton (i.e., unstabilized) iteration with step reduction line
search. This works very efficiently as long as $J$ is not ill-conditioned.
In a one-parameter problem with ill-conditioning, think of $J$ as a 
very small number so that we need to watch for a zero-divide or 0/0 
situation in computing $\delta$. Below we see `nls()` does poorly on this 
example problem. Note that `nlsr`
offers function `pnlm0()` to give a 1-line display of results of class `nls`,
and `pshort()` for `nlsr`.

@@

The singular values allow us to gauge how "nearly singular" the 
Jacobian is at the solution, and the ratio of the smallest to largest of the 
singular values is a simple but effective measure. In the above example, we note 
that the scaled logistic has the largest ("least singular") ratio.

The results from `nlsLM` and `gsl_nls` for the transformed model have a very 
large sum of squares, which may suggest that these programs have failed. 
Since `nls()`, `nlsLM()`, and `gsl_nls()` do not 
offer the singular values, we need to extract the Jacobian and compute its
singular values. The following script shows how to do this, using what is called
the `gradient` element in the returned solution for these solvers.

@@

More general termination tests can be used. `nlsr` has a **small sum of 
squares** test (**smallsstest**) that compares the latest evaluated sum of squared
(weighted) residuals to `e4` times the initial sum of squares, where
`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.

@@

Using the "gradient" attribute of the output of the Jacobian function to hold
this matrix lets us embed this attribute in the **residual** function as well,
so that the call to `nlfb()` can be made with the same name used for 
both residual and Jacobian function arguments. This programming trick saves a 
lot of trouble for the package developer, but it can be a nuisance for users 
trying to understand the code.

As far as we can understand the logic in `nls()`, the Jacobian computation during
parameter estimation is carried out within the called C-language program
and its wrapper R code function `numericDeriv()`, part of `./src/library/stats/R/nls.R`
in the R distribution source code. This is used to provide Jacobian information in 
the `nlsModel()` and `nlsModel.plinear()` functions, which are **not** exported for 
general use. `gsl_nls()` also appears to use `numericDeriv()`.

@@

`minpack.lm::nlsLM()` invokes `numericDeriv()` in its local 
version of `nlsModel()`, but it appears to use an internal approximate Jacobian 
code from 
the original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
can lead to different behaviour, though in our experience these differences are usually
very minor.

However, minor numerical differences can give annoying results with ill-conditioned
problems.

@@

`nls()` has an option `algorithm="plinear"` that allows some partially linear
models to be solved. The other tools, as far as we are aware, do not offer any
such capability. The `nlstac` package uses a different algorithm.

Within `nls()` itself we must, unfortunately, use
different specifications with different `algorithm` options.
The nonlinear model specifications are, of course, a development of linear ones.
The explicit model `y ~ a * x + b` does not work with the linear
modeling function `lm()`, which requires this model to be specified as `y ~ x`.
Within `nls()`, consider the following FOUR different specifications for the same 
problem, plus an intuitive choice, labelled `fm2a`, that does NOT work. 
In this failed attempt, putting the `Asym` parameter in the model causes the
`plinear` algorithm
to try to add another term to the model. We believe this is unfortunate, and would
like to see a consistent syntax. At the time of writing, we do
not foresee a resolution for this issue. In the example, we have NOT evaluated
the commands to save space.
